---
title: "WQ-Whit"
author: "Brie Sherow"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    df_print: paged
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-libraries, message=F, warning=F}
library(ggplot2) #graphing
library(ggthemes) #graphing templates
library(purrr) #looping
library(lubridate) #date manipulation
library(forcats) #working with factors
library(tidyverse) #manipulating data 
library(knitr) #rmarkdown functions
library(tidyr) #long & wide formats
library(reshape2) #melt function for wide and long formats
library(rio) #import xlsx with multiple sheets
library(kableExtra) #for displaying nice tables
library(ggstatsplot) #for finding outliers
library(DT) #excel-like searchable data tables
library(ggpubr) #grid prints for ggplot
library(formattable) #colour coding tables for grades
library(openxlsx) #fixing excel date formats

# Datatable good for anything that needs to be searchable, formattable for lovely design. Look into flextable, reactable, or gt as other option for formatting and colour-coding. 
# https://rstudio.github.io/DT/
#https://www.littlemissdata.com/blog/prettytables
#https://rfortherestofus.com/2019/11/how-to-make-beautiful-tables-in-r/
```

# Setup and references
```{r wq-index, message=F, warning=F}
options(scipen=999) # force non-scientific notation

# load WQ index
WQ_index <- import_list("data/WQ_Variables.xlsx", # GV, LOR, sites
            setclass = "tbl") # class is tibble, similar to dataframe

#site info
sites <- as.data.frame(WQ_index$sites)

# #lower limits of reporting for each partner - may be easier just to specify values in an equation when needed
# LOR <- as.data.frame(WQ_index$'WQ-LOR')
# # LOR wide to long
# LOR_long <- gather(LOR, Indicator, Value, PN:Turbidity, factor_key=TRUE)

# guideline values for failures
GV <- as.data.frame(WQ_index$GV_long) %>%
  mutate(Code = as.factor(Code),
         Name = as.factor(Name),
         Indicator = as.factor(Indicator),
         GV = as.numeric(Value),
         Stat = as.factor(Stat),
         Dir_Fail = as.factor(Failure)) %>%
  dplyr::select(-Failure, -Value)

```
Create vectors to distinguish data providers and zones.  Data providers include MMP, NQBP, and SIP.  Zones include northern, whitsundays, central, and southern.
```{r MMP-site-vectors, message=F, warning=F}
#create vectors of site codes
northern <- c("AP_AMB1", "AP_AMB2", "AP_AMB3", "AP_AMB4", "AP_AMB5")
whit <- c("WHI1", "WHI4", "WHI5")
central <- c("MKY_AMB1", "MKY_AMB2", "MKY_AMB3B", "MKY_AMB5", "MKY_AMB6B", "MKY_AMB8", "MKY_AMB10", "MKY_AMB12", "WHI6", "WHI7")
mmp <- c("WHI1", "WHI4", "WHI5", "WHI6", "WHI7")
nqbp <- c("AP_AMB1", "AP_AMB2", "AP_AMB3", "AP_AMB4", "AP_AMB5", "MKY_AMB1", "MKY_AMB2", "MKY_AMB3B", "MKY_AMB5", "MKY_AMB6B", "MKY_AMB8", "MKY_AMB10", "MKY_AMB12")
```

# Zone Level

## Whitsundays

### Raw data

****** TO DO: Check for duplicate entries, check outlier rules, add svg or table outputs to output folder

```{r import-data-nutrients, message=F, warning=F}
# load and set up MMP data  
      # load MMP nutrients
      nutrients_mmp <- read.csv("data/21_WQ_MMP_nutrients.csv")
            
      #remove unnecessary columns and set proper classes
      nutrients_mmp <- nutrients_mmp %>%
        filter(DEPTH_CODE == "D0") %>%
        mutate(Code = as.factor(SHORT_NAME),
               Name = as.factor(LOCATION_NAME), 
               Date = dmy(Date)) %>%
        select(Code, 
               Date,
               Name, SECCHI_DEPTH_m,
               NO2_uM, NO3_uM, PN_uM, PP_uM, 
               TSS_mg.L, Chl_ug.L)
```

```{r import-data-turbidity, message=F, warning=F}
# load and set up MMP data  
      # load MMP turbidity
      turbidity_mmp <- read.csv("data/21_WQ_MMP_turbidity.csv")

 #remove unnecessary columns and set proper classes
      turbidity_mmp <- turbidity_mmp %>%
        mutate(Code = as.factor(Code),
               Name = as.factor(Name), 
               Date = dmy(Date)) %>%
        select(Code,Name, Date, NTU_QA_AVG)
      
      turbidity_whit <- turbidity_mmp %>%
        filter(Code %in% whit)
      
#plot a time series for each site
turbidity_line_whit <- ggplot(turbidity_whit, aes(Date, NTU_QA_AVG)) + #x axis date, y axis NTU mean
  geom_line() + #line graph
  facet_wrap(~Name, scales = "free_y") + # separate plot for each site, axis range responsive to site
  theme(axis.text.x = element_text(angle = 45)) #angle axis text for clarity

ggsave("output/turbidity_line.png")
```

### Creating final data

```{r conversions-nutrients, message=F, warning=F}
      
      #converting units 
      nutrients_whit <- nutrients_mmp %>%
  filter(Code %in% whit) %>%
        mutate(NOx_uM = NO2_uM + NO3_uM,
               NOx_ug.l = NOx_uM*14,
               PN_ug.l = PN_uM*14,
               PP_ug.l = PP_uM*30.97) %>%
        select(Code:SECCHI_DEPTH_m, TSS_mg.L, 
               Chl_ug.L, NOx_ug.l, PN_ug.l, PP_ug.l) %>%
  rename(Secchi = SECCHI_DEPTH_m,
         TSS = TSS_mg.L,
         Chla = Chl_ug.L,
         NOx = NOx_ug.l,
         PN = PN_ug.l,
         PP = PP_ug.l)

#joining turbidity to nutrients

turbidity_whit_long <- turbidity_whit %>% 
  rename(Value = "NTU_QA_AVG") %>%
  mutate(Indicator = "NTU")

wq_whit <- nutrients_whit %>%
  gather(Indicator, Value, Secchi:PP) %>%
  rbind(turbidity_whit_long)
```

### Limits of Reporting

There were no LOR changes for MMP in 2020, so I'll have to reference other years to see what this looks like

### Outliers

```{r outliers, message=F, warning=F}      
outliers_whit <- wq_whit %>%
  group_by(Indicator) %>%
  mutate('5percentile' = quantile(Value, .05), #5th percentile for each indicator
         '95percentile' = quantile(Value, .95), #95th percentile for each indicator
         LowOutlier = ifelse(Value < quantile(Value,.05), "Low", "Pass"), #tagging values below 5th percentile
         HighOutlier = ifelse(Value > quantile(Value, .95), "High", "Pass"), #tagging values above 95th percentile
         Outlier = paste(LowOutlier, HighOutlier)) %>% #combine outlier columns
        mutate_at("Outlier", str_replace, "Pass", "") %>% #remove redundant entries
        mutate_if(is.character, str_trim) %>% #trim whitespace
        select(-LowOutlier, -HighOutlier) %>% #remove unnecessary columns
  filter(Outlier != "Pass")

datatable(outliers_whit)


```

### Clean data

```{r clean-data, message=F, warning=F}

datatable(nutrients_whit)
datatable(turbidity_whit)
```
### Guideline values 

```{r stats-table}
 # Create stats table 
      #Stats (site (name + code), indicator, n-samples, mean, min, 25%, median, 75%, max, direction of failure, stat, GV)
      
      #nutrients data wide to long, measuring against guideline values
      whit_stats <- wq_whit %>%
  left_join(GV, by = c("Code" = "Code", "Name" = "Name", "Indicator" = "Indicator")) %>%
        group_by(Code, Indicator) %>%
        mutate(Sample_Size = n(),
               Mean = mean(Value),
               Minimum = min(Value),
               '25th' = quantile(Value, probs = .25),
               Median = median(Value),
               '75th' = quantile(Value, probs = .75),
               Maximum = max(Value),
               Stat_val = ifelse(Stat %in% "mean", mean(Value), 
                                 ifelse(Stat %in% "median", median(Value),NA)),
               Failure = ifelse(Dir_Fail == "low" & Stat_val < GV, "too low",
                                ifelse(Dir_Fail == "high" & Stat_val > GV, "too high", "pass"))) %>%
        select(-Value, -Date)


datatable(whit_stats)
```
### Data values

```{r data-values}
#indicator avg data value
whit_values <- whit_stats %>%
        select(Code, Name, Indicator, Stat_val) %>%
        distinct() %>%
        pivot_wider(names_from = Indicator, 
                    values_from = Stat_val)

formattable(whit_values)
```

### Standardised scores

```{r standardised-scores}
# *** Secchi needs to be Stat_val/GV, all other indicators GV/Stat_val

#creating standardised scores by converting numeric values to grades from 0-100
whit_standardised <- whit_stats %>%
        select(Code, Name, Indicator, Stat_val, GV) %>%
        distinct() %>%
        mutate(standardised = case_when(
          Indicator %in% "Secchi" ~ifelse((log((Stat_val/GV),2))<=-1,-1,(ifelse((log((Stat_val/GV),2))>=1,1,(log((Stat_val/GV),2))))),
                                        TRUE ~ ifelse((log((GV/Stat_val),2))<=-1,-1,(ifelse((log((GV/Stat_val),2))>=1,1,(log((GV/Stat_val),2))))))) %>%
        select(-GV, -Stat_val) %>%
        pivot_wider(names_from = Indicator, 
                    values_from = standardised)

indicators <- c("Secchi", "TSS", "Chla", "NOx", "PN", "PP", "NTU")

datatable(whit_standardised, rownames = FALSE) %>%
  formatStyle(columns = indicators, 
              background = styleInterval(c(-.66, -.33, 0, .51)-1e-6, c("#FF0000", "#FFC000", "#FFFF00", "#92D050", "#00B050")))
```
### Category scores

```{r category-scores}
#Averaging values to condense categories
whit_categories <- whit_standardised %>%
  mutate(Nutrients = mean(c(NOx, PN, PP)),
         Chla = Chla,
         Clarity = mean(c(Secchi,TSS, NTU))) %>%
  select(Name, Code, Nutrients, Chla, Clarity)

indicators <- c("Chla", "Nutrients", "Clarity")

datatable(whit_categories, rownames = FALSE) %>%
  formatStyle(columns = indicators, 
              background = styleInterval(c(-.66, -.33, 0, .51)-1e-6, c("#FF0000", "#FFC000", "#FFFF00", "#92D050", "#00B050")))
```
### Zone scores

```{r zone-scores}
whit_zone_score <- colMeans(whit_categories[c("Nutrients", "Chla", "Clarity")]) %>%
  t() %>%
  as.data.frame() %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, everything())

indicators <- c("Chla", "Nutrients", "Clarity")

datatable(whit_zone_score, rownames = FALSE) %>%
  formatStyle(columns = indicators, 
              background = styleInterval(c(-.66, -.33, 0, .51)-1e-6, c("#FF0000", "#FFC000", "#FFFF00", "#92D050", "#00B050")))
```

### Zone grade

***** Very suspicious that Chl-a has gone from moderate in 2020 to very good in 2021...  need to check these numbers.
```{r zone-grade}
whit_zone_grade <- whit_zone_score %>%
  transmute(Zone = "Whitsunday",
          Nutrients = round(40.9- (19.9 - ((Nutrients -(-0.66)) * (19.9/0.32)))),
          Chla = round(60.9- (19.9 - ((Chla -(-0.33)) * (19.9/0.32)))),
          Clarity = round(40.9- (19.9 - ((Clarity -(-0.66)) * (19.9/0.32))))) %>%
  mutate('Water Quality' = round(mean(Nutrients:Clarity)))

indicators <- c("Chla", "Nutrients", "Clarity", "Water Quality")

datatable(whit_zone_grade, rownames = FALSE) %>%
  formatStyle(columns = indicators, 
              background = styleInterval(c(21, 41, 61, 81)-1e-6, c("#FF0000", "#FFC000", "#FFFF00", "#92D050", "#00B050")))
```
# Indicator Level

## TSS

```{r outliers-TSS}
outliers <- boxplot(nutrients_whit$TSS)$out 

TSS_whit_outliers <- nutrients_whit %>%
  select(Code, Name, Date, TSS) %>%
  mutate(Outlier = ifelse(TSS %in% outliers, "TRUE", "FALSE"))

formattable(TSS_whit_outliers)
```
```{r quantile-TSS}

TSS_quantile <- nutrients_whit %>%
  select(Code, Name, Date, TSS) %>%
  mutate(Min = min(TSS), #minimum
         Q1 = quantile(TSS, .25),
         Median = median(TSS),
         Q3 = quantile(TSS, .75),
         Max = max(TSS),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -TSS) %>%
        distinct() %>%
        t()

formattable(TSS_quantile) %>%
  kable("html", caption = "TSS: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F)


```
```{r boxplot-TSS}
boxplot_TSS <- nutrients_whit %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, TSS) %>%
  ggplot() +
    geom_boxplot(aes(x=Zone, y = TSS))

ggsave("output/boxplot_TSS.png")
```

## Chlorophyll-a

```{r outliers-Chla}
outliers <- boxplot(nutrients_whit$Chla)$out 

Chla_whit_outliers <- nutrients_whit %>%
  select(Code, Name, Date, Chla) %>%
  mutate(Outlier = ifelse(Chla %in% outliers, "TRUE", "FALSE"))

formattable(Chla_whit_outliers)
```

```{r quantile-Chla}

Chla_quantile <- nutrients_whit %>%
  select(Code, Name, Date, Chla) %>%
  mutate(Min = min(Chla), #minimum
         Q1 = quantile(Chla, .25),
         Median = median(Chla),
         Q3 = quantile(Chla, .75),
         Max = max(Chla),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -Chla) %>%
        distinct() %>%
        t()

formattable(Chla_quantile) %>%
  kable("html", caption = "Chlorophyll-a: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F)
```
```{r boxplot-Chla}
boxplot_chla <- nutrients_whit %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, Chla) %>%
  ggplot() +
    geom_boxplot(aes(x=Zone, y = Chla))

ggsave("output/boxplot_Chla.png")
```

## NOx

```{r outliers-NOx}
outliers <- boxplot(nutrients_whit$NOx)$out 

NOx_whit_outliers <- nutrients_whit %>%
  select(Code, Name, Date, NOx) %>%
  mutate(Outlier = ifelse(NOx %in% outliers, "TRUE", "FALSE"))

formattable(NOx_whit_outliers)
```

```{r quantile-NOx}

NOx_quantile <- nutrients_whit %>%
  select(Code, Name, Date, NOx) %>%
  mutate(Min = min(NOx), #minimum
         Q1 = quantile(NOx, .25),
         Median = median(NOx),
         Q3 = quantile(NOx, .75),
         Max = max(NOx),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -NOx) %>%
        distinct() %>%
        t()

formattable(NOx_quantile) %>%
  kable("html", caption = "NOx: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F) %>%
  save_kable("output/NOx_quantile.pdf")
```
```{r boxplot-NOx}
boxplot_NOx <- nutrients_whit %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, NOx) %>%
  ggplot() +
    geom_boxplot(aes(x=Zone, y = NOx))

ggsave("output/boxplot_NOx.png")
```

## Secchi


```{r outliers-Secchi}
outliers <- boxplot(nutrients_whit$Secchi)$out 

Secchi_whit_outliers <- nutrients_whit %>%
  select(Code, Name, Date, Secchi) %>%
  mutate(Outlier = ifelse(Secchi %in% outliers, "TRUE", "FALSE"))

formattable(Secchi_whit_outliers) %>%
  save_kable("output/Secchi_whit_outliers.pdf")
```

```{r quantile-Secchi}

Secchi_quantile <- nutrients_whit %>%
  select(Code, Name, Date, Secchi) %>%
  mutate(Min = min(Secchi), #minimum
         Q1 = quantile(Secchi, .25),
         Median = median(Secchi),
         Q3 = quantile(Secchi, .75),
         Max = max(Secchi),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -Secchi) %>%
        distinct() %>%
        t()

formattable(Secchi_quantile) %>%
  kable("html", caption = "Secchi: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F) %>%
  save_kable("output/Secchi_quantile.pdf")
```
```{r boxplot-Secchi}
boxplot_secchi <- nutrients_whit %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, Secchi) %>%
  ggplot() +
    geom_boxplot(aes(x=Zone, y = Secchi))

ggsave("output/boxplot_Secchi.png")
```


## PN

```{r outliers-PN}
outliers <- boxplot(nutrients_whit$PN)$out 

PN_whit_outliers <- nutrients_whit %>%
  select(Code, Name, Date, PN) %>%
  mutate(Outlier = ifelse(PN %in% outliers, "TRUE", "FALSE"))

#*** Need to create somehting in cases where there's no outliers
formattable(PN_whit_outliers)
```

```{r quantile-PN}

PN_quantile <- nutrients_whit %>%
  select(Code, Name, Date, PN) %>%
  mutate(Min = min(PN), #minimum
         Q1 = quantile(PN, .25),
         Median = median(PN),
         Q3 = quantile(PN, .75),
         Max = max(PN),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -PN) %>%
        distinct() %>%
        t()

formattable(PN_quantile) %>%
  kable("html", caption = "PN: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F) %>%
  save_kable("output/PN_quantile.pdf")
```
```{r boxplot-PN}
boxplot_PN <- nutrients_whit %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, PN) %>%
  ggplot() +
    geom_boxplot(aes(x=Zone, y = PN))

ggsave("output/boxplot_PN.png")
```
## PP

```{r outliers-PP}
outliers <- boxplot(nutrients_whit$PP)$out 

PP_whit_outliers <- nutrients_whit %>%
  select(Code, Name, Date, PP) %>%
  mutate(Outlier = ifelse(PP %in% outliers, "TRUE", "FALSE"))

formattable(PP_whit_outliers)
```

```{r quantile-PP}

PP_quantile <- nutrients_whit %>%
  select(Code, Name, Date, PP) %>%
  mutate(Min = min(PP), #minimum
         Q1 = quantile(PP, .25),
         Median = median(PP),
         Q3 = quantile(PP, .75),
         Max = max(PP),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -PP) %>%
        distinct() %>%
        t()

formattable(PP_quantile) %>%
  kable("html", caption = "PP: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F) %>%
  save_kable("output/PP_quantile.pdf")
```
```{r boxplot-PP}
boxplot_PP <- nutrients_whit %>%
  mutate(Zone = "Whitsunday") %>%
  select(Zone, PP) %>%
  ggplot() +
    geom_boxplot(aes(x=Zone, y = PP))

ggsave("output/boxplot_PP.png")
```

## NTU

```{r outliers-NTU}

outliers <- boxplot(turbidity_whit$NTU_QA_AVG)$out 

NTU_whit_outliers <- turbidity_whit %>%
  rename(NTU = NTU_QA_AVG) %>%
  mutate(Outlier = ifelse(NTU %in% outliers, "TRUE", "FALSE"))

# Table of outliers
datatable(NTU_whit_outliers)
```

```{r quantile-NTU}

NTU_quantile <- turbidity_whit %>%
  rename(NTU = NTU_QA_AVG) %>%
  mutate(Min = min(NTU), #minimum
         Q1 = quantile(NTU, .25),
         Median = median(NTU),
         Q3 = quantile(NTU, .75),
         Max = max(NTU),
         IQR = Q3 - Q1,
         LowWhisker = Q1 - 1.5*IQR,
         HighWhisker = Q3 + 1.5*IQR) %>%
        select(-Code, -Name, -Date, -NTU) %>%
        distinct() %>%
        t()

formattable(NTU_quantile) %>%
  kable("html", caption = "NTU: quantile by zone", escape = F, col.names = "Whitsundays")%>% 
  kable_styling(full_width = F)%>%
  save_kable("output/NOx_quantile.pdf")
```
```{r boxplot-NTU}
sample_size <- turbidity_whit %>% group_by(Code) %>% summarize(num=n())

 #Boxplots by site
    boxplot_NTU <-  turbidity_whit %>%
       rename(NTU = NTU_QA_AVG) %>%
        left_join(sample_size, by = "Code") %>%
        mutate(Site = paste0(Name, "\n", "n=", num)) %>% #print sample size next to site name
      ggplot( aes(x=Site, y=NTU)) + #calling NTU mean as y and site as x
        # geom_boxplot(varwidth = TRUE) #create boxplot with width determined by sample size
        geom_violin(width = 1.4, alpha=0.5) + #violin plot more accurately depicts distribution and sample size
        geom_boxplot(width=0.1) + #depict boxplot inside of the violin plot
        geom_jitter(color="grey", size=0.7, alpha=0.4) #add all points to visualise distribution and sample size

     ggsave("output/boxplot_NTU.png")
```