---
title: "WQ"
author: "Brie Sherow"
date: "13/10/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### Reference links
# R-Markdown: https://bookdown.org/yihui/rmarkdown-cookbook/word-template.html

# GitHub: https://drive.google.com/drive/folders/1HnqhDJRCVebfGyusHI01-LudLfojG9vM
```


```{r load-libraries, message=F, warning=F}
library(ggplot2) #graphing
library(ggthemes) #graphing templates
library(purrr) #looping
library(lubridate) #date manipulation
library(forcats) #working with factors
library(tidyverse) #manipulating data (includes heaps of libraries)
library(knitr) #rmarkdown functions
library(tidyr) #long & wide formats
library(reshape2) #melt function for wide and long formats
library(rio) #import xlsx with multiple sheets
library(kableExtra) #for displaying nice tables
library(ggstatsplot) #for finding outliers
library(DT) #excel-like searchable data tables
```
1. Data quality checks for missing values, LORs, and outliers in excel. Can outliers or missing values be explained by an event or are there mistakes? Flag and comment in excel, go back to data source to check when needed. LOR/LOD check (mainly SIP) transform values too low for detection, check limits and calculations (LOR * 0.5). Import this corrected raw data into R.

# WQ Index
```{r}
# load WQ index
WQ_index <- import_list("data/WQ_Variables.xlsx", # GV, LOR, sites
            setclass = "tbl") # class is tibble, similar to dataframe

#site info
sites <- as.data.frame(WQ_index$sites)

# #lower limits of reporting for each partner
# LOR <- as.data.frame(WQ_index$'WQ-LOR')
# # LOR wide to long
#         # *** need to add column to describe whether the value is mean, median, etc
# LOR_long <- gather(LOR, Indicator, Value, PN:Turbidity, factor_key=TRUE)

# guideline values for failures
GV <- as.data.frame(WQ_index$'WQ-GV')

```

2. Raw data in excel (Code, Name, Date, each indicator its own column, LOR and outliers highlighted). 

2) Final data (same as raw but with outliers and LOR corrected). Import this final data into R.

*** Head should look like this: ACTION to put a jpeg of data head***

# Northern ====================================

## NQBP data
```{r NQBP-site-vectors}
#create vectors of site codes
northern <- c("AP_AMB1", "AP_AMB2", "AP_AMB3", "AP_AMB4", "AP_AMB5")
central<- c("MKY_AMB1", "MKY_AMB2", "MKY_AMB3B", "MKY_AMB5", "MKY_AMB6B", "MKY_AMB8", "MKY_AMB10", "MKY_AMB12", "WHI6", "WHI7")
#note that all central sites are included in this vector, not jsut NQBP
NQBP_sites <- c(northern, "MKY_AMB1", "MKY_AMB2", "MKY_AMB3B", "MKY_AMB5", "MKY_AMB6B", "MKY_AMB8", "MKY_AMB10", "MKY_AMB12")
```

    
## Join nutrients and secchi

```{r import-data, message=F, warning=F}
  
# load nutrients and secchi
WQ_NQBP <- import_list("data/WQ_NQBP_20_nutrients_LORs-changed.xlsx", # all sites (each site is a tab)
            setclass = "tbl") # class is tibble, similar to dataframe
      
      #import NQBP nutrients
      nutrients <- WQ_NQBP$Nutrients %>%
        mutate(Code = as.factor(Site_code), #recoding indicators to numeric, factor, and date as needed
               Name = as.factor(Site_name),
               PN = as.numeric(PN),
               Date=ymd(Sample_date),
               Chla = Chl_a) %>%
        dplyr::select(Code, Name, Date, TSS, NOx, PN, PP, Chla) #select only relevant indicators
      
      
      #import NQBP secchi
      secchi <- WQ_NQBP$Physiochem %>%
        mutate(Code = as.factor(Site_code), #recoding indicators to numeric, factor, and date as needed
               Name = as.factor(Site_name),
               Date=ymd(Date),
               Season = as.factor(Season),
               Secchi = as.numeric(Secchi)) %>%
        dplyr::select(Code, Name, Date, Season, Secchi) %>% #select indicators
        distinct()
      
      
#join secchi to wq 
# ***check that both have the same number of observations
      WQN_wide <- nutrients %>% #join to secchi so as not to lose the row of observations that are missing in nutrients
        full_join(secchi, by = c("Code" = "Code", "Date" = "Date")) %>% #join by two fields
        mutate(Name = Name.x) %>%
        dplyr::select(Code, Name, Date, TSS, NOx, PN, PP, Chla, Secchi, Season) %>% #select only relevant indicators
        filter(Code %in% NQBP_sites) %>%
        mutate(zone = 
                 ifelse(Code %in% northern, "Northern", "Central"),
               supplier = "NQBP") %>%
        distinct()
  
# # data wide to long
#         # *** need to add column to describe whether the value is mean, median, etc
# WQ.N <- gather(WQN_wide, Indicator, Value, TSS:Secchi, factor_key=TRUE)

```

    
## NTU checks

2. Make plots of NTU

```{r NTU-plots, message=F, warning=F}

#import NTU raw data
ntu_raw <- import_list("data/WQ_NQBP_20_NTUe.xlsx", # all sites (each site is a tab)
            setclass = "tbl", # class is tibble, similar to dataframe
            rbind = TRUE) # joins all xlsx sheets into one table
      
#filter for moitoring sites
ntu <- as.data.frame(ntu_raw) %>% #change from tibble to dataframe
        filter(Site_code %in% NQBP_sites) %>% #select report card sites
        mutate(NTUe_mean = as.numeric(NTUe_mean), #change mean and sd from character to numeric
               NTUe_sd = as.numeric(NTUe_sd),
               zone = ifelse(Site_code %in% northern, "Northern", "Central"),
               supplier = "NQBP") %>%
        distinct()

#plot a time series for each site
ggplot(ntu, aes(Date, NTUe_mean)) + #x axis date, y axis NTUe mean
  geom_line() + #line graph
  facet_wrap(~Site_name, scales = "free_y") + # separate plot for each site, axis range responsive to site
  theme(axis.text.x = element_text(angle = 45)) #angle axis text for clarity
```

## WQ checks
    
  2. Boxplots, quartile valus, and outlier tables for each indicator
  
```{r boxplots-indicators, message=F, warning=F}
#WQ indicators
      #create a vector of indicator names
      indicators <- WQN_wide %>%
        select_if(is.numeric) %>% # select all columns with numeric data
        names() %>% #identifies the names of each column
        set_names() #create a vector out of the column names


      #Quartile values output tables
              QV <- map(indicators, function(x){
                
                #create temp dataframe
                temp_df <- WQN_wide %>%
                  dplyr::select(Code, Name, Date, paste0(x), zone) %>% #the paste0(x) is iterating through the indicators vector
                  na.omit() #removing NA values
                
                #quartile values summary
                summary <- temp_df %>%
                  select(x) %>% #this is iterating through the 5 indicators
                  summary() #quartile values summary of the selected indicator
                
                return(summary)
              })
              
              QV %>% #prints table
                 kable("html") %>%
                kable_styling(full_width=F) #make the table look nice
     
      #oulier output tables
              indicator_outliers <- map(indicators, function(x){
                
                #create temp dataframe
                temp_df <- WQN_wide %>%
                dplyr::select(Code, Name, Date, paste0(x), zone) %>% #the paste0(x) is iterating through the indicators vector
                na.omit() #remove NA values
                
                #creates vector of outliers
                outliers <- boxplot.stats(temp_df[[x]])
              
                #outlier output table
                temp_out <- temp_df %>%
                mutate(Outlier = ifelse(x %in% outliers, "TRUE", "FALSE")) %>% #if value matches outlier vector, assign true, if not assign false
                select(Date, Code, x, Outlier) #select columns to export
                
                return(temp_out)
        
                })
              
              
                kable(indicator_outliers, "simple", 
                      booktabs=T, valign = 't',
                      align = "lccr") %>%
                kable_styling() #makes it look nice
                
                datatable(data=indicator_outliers$PP)


   #Boxplots by site   
      boxplots <- map(indicators, #setting the items to iterate, in this case our vector of five indicators
                      ~ggplot(data = WQN_wide) + #setting the dataframe to use
                       geom_boxplot(aes(x = Name, y = .data[[.x]]) ) + #site name x, indicator y
                       labs(y = .x) + #labels the y axis with indicator vector 
                       theme(axis.text.x = element_text(angle = 45)))
      
      boxplots #prints boxplots
      
      ### Reference links
      # https://stackoverflow.com/questions/58204452/printing-ggplot-with-purrr-map
```
## After final data


```{r boxplots-NTU, message=F, warning=F}      
# NTU
  #Summary output tables
      #remove NA values
      ntuN <- ntuN[!is.na(ntuN$NTUe_mean), ] #create a dataframe removing all instances where NTU mean is NA
      
      #create a vector of outlier values
      outliers <- boxplot(ntuN$NTUe_mean)$out 
      
      #outlier output table
        ntuN_outliers <- ntuN %>%
        mutate(Outlier = ifelse(NTUe_mean %in% outliers, "TRUE", "FALSE")) %>%
        select(Date, Site_code, NTUe_mean, Outlier)
        
        kable(ntuN_outliers[1:10, ],"html") %>% #first ten entries (this is a long table)
          kable_styling(full_width=F) #make the table look nice
      
      #quartile values output table
      ntuN %>%
        select(NTUe_mean) %>%
        summary() %>%
        kable("html") %>% kable_styling(full_width=F) #make the table look nice
      
  #boxplots per site
      #check to see that sample sizes are the same for each site
      sample_size <- ntuN %>% group_by(Site_name) %>% summarize(num=n()) #number of entries for each site
      sample_size %>% #print a table that shows the number of entries for each site
        kable("html", col.names = c("Site", "NTU sample size")) %>% kable_styling(full_width=F) #make the table look nice
      #***Holbourne Island has the lowest sample size at 189 and Euri Creek is the highest at 234
      
      #Boxplots by site
     ntu_boxplot <-  ntuN %>%
        left_join(sample_size, by = "Site_name") %>%
        mutate(site = paste0(Site_name, "\n", "n=", num)) %>% #print sample size next to site name
      ggplot( aes(x=site, y=NTUe_mean)) + #calling NTU mean as y and site as x
        # geom_boxplot(varwidth = TRUE) #create boxplot with width determined by sample size
        geom_violin(width = 1.4, alpha=0.5) + #violin plot more accurately depicts distribution and sample size
        geom_boxplot(width=0.1) + #depict boxplot inside of the violin plot
        geom_jitter(color="grey", size=0.7, alpha=0.4) #add all points to visualise distribution and sample size

     ntu_boxplot #prints plot
      
      ###Reference links
      #https://rstudio-pubs-static.s3.amazonaws.com/116317_e6922e81e72e4e3f83995485ce686c14.html#/ (mutate with ifelse statements)
      #https://www.data-to-viz.com/caveat/boxplot.html
```

## Output tables

3) Stats (site (name + code), indicator, n-samples, mean, min, 25%, median, 75%, max, direction of failure, stat, GV)

## Indicator tables

1) use stats table to get annual mean or median for each site

2) calc unstandardised indicator score - if the log of GV value mean or median / real value is less than 1, =1, or 2 - basically calculating whether its higher or lower than guideline value

3) Avg of indicators to form indicator category (need to have 60% of indicators to grade)

4) Indicator zone level (avg of sties)

5) standardised score (join indicator zone level score to WQ index to determine Condition and resulting equation to calc GBR range)

6) avg of nutrients, chl-a, clarity, and pesticides to get WQ grade, need 60% (3/4) to write results.

# Whitsundays=======================
MMP and BP data
```{r MMP-site-vectors}
#create vectors of site codes
whit <- c("WHI1", "WHI4", "WHI5")
central <- c("MKY_AMB1", "MKY_AMB2", "MKY_AMB3B", "MKY_AMB5", "MKY_AMB6B", "MKY_AMB8", "MKY_AMB10", "MKY_AMB12", "WHI6", "WHI7")
mmp_sites <- c("WHI1", "WHI4", "WHI5", "WHI6", "WHI7")
```


  1. MMP conversions
    calculate NOX, convert units for NOX, PN, PP
```{r import-data, message=F, warning=F}
# load and set up MMP data  
      # load MMP nutrients
      nutrients_mmp <- read.csv("data/21_WQ_MMP_nutrients.csv") 
            
      #remove unnecessary columns and classify type
      nutrients_mmp <- nutrients_mmp %>%
        mutate(Code = as.factor(SHORT_NAME),
               Name = as.factor(LOCATION_NAME),
               Date = dmy(Date),
               Depth = as.factor(DEPTH_CODE)) %>%
        select(Depth, Code, Date, Name, Secchi_depth_m, 
               NO2_uM, NO3_uM, PN_SHIM_uM, PP_uM, 
               TSS_mg.l, Chl_ug.l) 
      
      #conversions
      nutrients_mmp <- nutrients_mmp %>%
        mutate(NOx_uM = NO2_uM + NO3_uM,
               NOx_ug.l = NOx_uM*14,
               PN_ug.l = PN_SHIM_uM*14,
               PP_ug.l = PP_uM*30.97) %>%
        select(Depth:Secchi_depth_m, TSS_mg.l, 
               Chl_ug.l, NOx_ug.l, PN_ug.l, PP_ug.l)
      
# Comparisons between surface and depth samples   
        # create wide df to calc across depths
      mmp_nut_wide <- nutrients_mmp %>%
        pivot_wider(names_from = Depth, 
                    values_from = TSS_mg.l:PP_ug.l) %>%
        mutate(TSS_avg = 
                 rowMeans(select(.,TSS_mg.l_D0: TSS_mg.l_D1)),
               Chl_avg = 
                 rowMeans(select(.,Chl_ug.l_D0: Chl_ug.l_D1)),
               NOx_avg =
                 rowMeans(select(.,NOx_ug.l_D0: NOx_ug.l_D1)),
               PN_avg =
                 rowMeans(select(.,PN_ug.l_D0: PN_ug.l_D1)),
               PP_avg =
                 rowMeans(select(.,PP_ug.l_D0: PP_ug.l_D1))) %>%
        select(Code, Date, TSS_avg:PP_avg)
      

      
nutrients_mmp <- nutrients_mmp %>%
  filter(Depth == "D0") %>%
  left_join(mmp_nut_wide, by = c("Code", "Date"))

```
# Central=======================
MMP and NQBP data

  1.MMP conversions
    calculate NOX, convert units for NOX, PN, PP

# Southern=======================
SIP data